\chapter{Analytical phase diversity algorithm} 
\label{ch:ourPD}

After having tested and used the ONERA algorithm, we can pass to the main goal of this work, the development of our phase diversity algorithm. In this chapter, we will describe the algorithm and its implementation. And we will present the result of its testing with simulated PSFs. We will present also an approach, that could be used on a telescope equipped with a deformable mirror, increasing the validity domain of the method. And finally we will compare our algorithm with the ONERA algorithm.

\section{Analytical algorithm}
\label{sec:AnAlgo}

\subsection{Algorithm description}
\label{subsec:ANalgoDesc}

This algorithm uses an analytical approach, developed by Dr. Laurent Jolissaint, to retrieve the phase of the wavefront \textbf{induced by a known point source object}. We assume the object known, because we want to correct for the static aberrations present in the optical system to the scientific detector and thus we use a point source to illuminate the optical system, either a star or a laser.

As we have seen in section \ref{sec:ImSystem}, the PSF of an optical system correspond to the image it gives of a point source,

\begin{equation}
PSF(x,y) = \frac{1}{S_p^2}|\left[\mathcal{F}\left\lbrace P(\xi,\eta)A(\xi,\eta)e^{-j\phi(\xi,\eta)} \right\rbrace\right](x,y)|^2,
\label{eqt:PSF}
\end{equation}

where $P(\xi,\eta)$ is the exit pupil function, $A(\xi,\eta)$ is the amplitude of the wave through the exit pupil, $\phi(\xi,\eta)$ is the phase of the wavefront and $S_p$ is the exit pupil surface. In the following we will omit the coordinates to simplify the notation. The unit of the PSF is directly the Strehl ratio. Under the assumption that we have weak aberrations, we can expand the exponential term,

\begin{equation}
exp(-j\phi)\approx 1 - j\phi - \frac{\phi^2}{2} + O(\phi^3),
\label{eqt:expansionPhase}
\end{equation}

replacing the exponential by its expansion in eqt.\eqref{eqt:PSF} leads to,

\begin{equation}
S_p^2 PSF \cong |\mathcal{F}\left\lbrace PA (1-j\phi-\frac{\phi^2}{2}) \right\rbrace|^2
\label{eqt:PSFwthPhaseExpand}
\end{equation}

Developing eqt. \eqref{eqt:PSFwthPhaseExpand}, keeping only the terms up to the second order, assuming that the amplitude through the pupil $A(\xi,\eta)$ is constant and unitary since we have a point source object and using the well known complex relations,

\begin{align}
a + a^* &= 2 \Re \lbrace a \rbrace \nonumber \\
a - a^* &= 2j \Im \lbrace a \rbrace, \nonumber
\end{align}

we obtain the following relation,

\begin{equation}
S_p^2 PSF \cong |\widetilde{P}|^2 + |\widetilde{P\phi}|^2 + 2\Im\lbrace \widetilde{P^*}\widetilde{P \phi}\rbrace - 2\Re\lbrace \widetilde{P^*}\widetilde{P \phi^2}\rbrace
\label{eqt:devPSFwthPhaseExpand}
\end{equation}

Defining $\Delta PSF$ as the difference between eqt. \eqref{eqt:devPSFwthPhaseExpand} for an arbitrary optical system and its perfect equivalent, we obtain the following expression,

\begin{equation}
\Delta PSF = S_p^2 PSF - S_p^2 PSF_{perfect} = |\widetilde{P\phi}|^2 + 2\Im\lbrace \widetilde{P^*}\widetilde{P \phi}\rbrace - 2\Re\lbrace \widetilde{P^*}\widetilde{P \phi^2}\rbrace,
\label{eqt:DeltaPSF}
\end{equation}

where $S_p^2 PSF_{perfect}$ is equal for a equivalent perfect system with the same pupil to $|\widetilde{P}|^2$. One can decompose $\phi$ into its even and odd phase, $\psi$ and $\gamma$ respectively,

\begin{equation}
\phi = \psi + \gamma
\label{eqt:Phidecomposed}
\end{equation}

Developing eqt. \eqref{eqt:DeltaPSF} after replacing $\phi$ by its decomposition and using the properties of the Fourier transform of real and purely even or odd functions, we get the following expression,

\begin{equation}
\Delta PSF = |\widetilde{P\psi}|^2 + |\widetilde{P\gamma}|^2 + 2\Im\lbrace \widetilde{P^*}\widetilde{P \gamma}\rbrace - \Re\lbrace \widetilde{P^*}\widetilde{P \psi^2}\rbrace- \Re\lbrace \widetilde{P^*}\widetilde{P \gamma^2}\rbrace
\label{eqt:DeltaPSFdeveloped}
\end{equation}

We can decompose $\Delta PSF$ into its even and odd components,

\begin{align}
\Delta PSF_{even} &= |\widetilde{P\psi}|^2 + |\widetilde{P\gamma}|^2 - \Re\lbrace \widetilde{P^*}\widetilde{P \psi^2}\rbrace- \Re\lbrace \widetilde{P^*}\widetilde{P \gamma^2}\rbrace, \label{eqt:DeltaPSFeven}\\
\Delta PSF_{odd} &= 2\Im\lbrace \widetilde{P^*}\widetilde{P \gamma}\rbrace \label{eqt:DEltaPSFodd},
\end{align}

This equation system shows that we can retrieve the odd part of the phase easily with eqt. \eqref{eqt:DEltaPSFodd}. But eqt. \eqref{eqt:DeltaPSFeven} clearly reveals the indetermination of the phase retrieval with only one image, as the sign of the even part of $\phi$ can not be determine. In order to raise this indetermination, as exposed in section \ref{sec:principle}, we need to introduce a phase diversity $\delta\phi$. We can modify the pupil function $P$ in order to take into account this introduced diversity,

\begin{equation}
P_{\delta} \equiv P e^{-j\delta\phi} = P(cos(\delta\phi)-jsin(\delta\phi)) = P(C-iS) 
\label{eqt:pupilDeltaphi}
\end{equation}

The expression of $\Delta PSF_{\delta\phi}$, which is the $\Delta PSF$ at the defocus plane, is found by replacing $P$ by $P_{\delta}$ in eqt. \eqref{eqt:DeltaPSFdeveloped}, we give directly the expressions of the even and odd components by taking into account that the phase is only define on the pupil ($P\phi=\phi$) to simplify the reading,

\begin{align}
\Delta PSF_{\delta\phi, even} &= |\widetilde{C\psi}|^2 + |\widetilde{C\gamma}|^2 +|\widetilde{S\psi}|^2 + |\widetilde{S\gamma}|^2 -2\widetilde{PC}^*\widetilde{S\psi}+2\widetilde{PS}^*\widetilde{C\psi} \nonumber\\
&-\widetilde{PC}^*\widetilde{C\psi^2}-\widetilde{PC}^*\widetilde{C\gamma^2}-\widetilde{PS}^*\widetilde{S\psi^2}-\widetilde{PS}^*\widetilde{S\gamma^2} \label{eqt:DeltaPSFevenDef}\\
\Delta PSF_{\delta\phi, odd} &= 2\widetilde{C\psi}^*\Im\lbrace\widetilde{S\gamma}\rbrace+2\Im\lbrace\widetilde{C\gamma}^*\rbrace\widetilde{S\psi}+2\widetilde{PC}^*\Im\lbrace\widetilde{C\gamma}\rbrace+2\widetilde{PS}^*\Im\lbrace\widetilde{S\gamma}\rbrace \nonumber\\
&+2j\widetilde{PC}^*\widetilde{S\psi\gamma}-2j\widetilde{PS}^*\widetilde{C\psi\gamma}.\label{eqt:DeltaPSFoddDef}
\end{align}

Eqt. \eqref{eqt:DEltaPSFodd}, eqt. \eqref{eqt:DeltaPSFevenDef} and eqt. \eqref{eqt:DeltaPSFoddDef} allow to retrieve the complete phase of the optical system under the assumption of weak aberrations. The retrieval numerical method uses the decomposition of the even and odd part of the phase on the Zernike polynomials,

\begin{align}
\psi &= \sum\limits_{js\ even} a_j Z_j \label{eqt:evenPhaseDecomp}\\
\gamma &= \sum\limits_{js\ odd} a_j Z_j. \label{eqt:oddPhaseDecomp}
\end{align}

This allows to have a linear system of equations with respect to the Zernike coefficient $a_j$ using eqts. \eqref{eqt:DEltaPSFodd} and \eqref{eqt:DeltaPSFoddDef}, but a problem arises as we tried to retrieve the phase of a purely even phase. The equations gave us an odd phase part equals to zero but also an even phase part equal to zero. This comes from the fact that in eqt. \eqref{eqt:DeltaPSFoddDef}, each term is multiplied by the odd phase component. And we could not use eqts. \eqref{eqt:DeltaPSFeven} or \eqref{eqt:DeltaPSFevenDef}, due to the squared modulus of the even and odd phase component, which rendered our system of of equation non-linear.

One way to get around this issue is to add another diversity, which is equal in amplitude to the first one but its inverse. So we have two diversity given by,

\begin{equation}
\delta\phi_+ = \delta\phi \mathrm{\ and \ } \delta\phi_- = -\delta\phi
\label{eqt:diversities}
\end{equation}

Using the new diversity, eqt. \eqref{eqt:DEltaPSFodd} allows to determine the odd part of the phase as before, but now for the even part of the phase we compute the difference between $\Delta PSF_{\delta\phi_+,even}$ and $\Delta PSF_{\delta\phi_-,even}$. This gives us the following system of equation,

\begin{align}
\Delta PSF_{odd} &= 2\Im\lbrace \widetilde{P^*}\widetilde{P \gamma}\rbrace \\
\Delta PSF_{\delta\phi_+, even}-\Delta PSF_{\delta\phi_-, even} &= -4\widetilde{PC}^*\widetilde{S\psi} +4\widetilde{PS}^*\widetilde{C\psi},
\end{align}

which we can use to determine the complete phase of the wavefront. We can rewrite it using the decomposition of $\psi$ and $\gamma$ on the Zernike basis,

\begin{align}
\Delta PSF_{odd} &= \sum\limits_{js\ odd} a_j 2\Im\lbrace \widetilde{P^*}\widetilde{P Z_j}\rbrace \label{eqt:DeltaPSFoddonZernike}\\
\Delta PSF_{\delta\phi_+, even}-\Delta PSF_{\delta\phi_-, even} &= \sum\limits_{js\ even} a_j \lbrace-4\widetilde{PC}^*\widetilde{SZ_j} +4\widetilde{PS}^*\widetilde{CZ_j}\rbrace.\label{eqt:DeltaPSF+-DeltaPSF-evenonZernike}
\end{align}

To solve these two equations and find the $a_j$'s, we use a linear regression method. We can rewrite the equations under a vectorial form to clarify the equations by flattening the images,

\begin{align}
\overrightarrow{\Delta PSF}_{odd} &= \underline{Z}_f\vec{a}_{odd}  \label{eqt:DeltaPSFoddonZernikeMAtrix}\\
\overrightarrow{\Delta PSF}_{\delta\phi_+, even}-\overrightarrow{\Delta PSF}_{\delta\phi_-, even} &= \underline{Z}_d \vec{a}_{even}.\label{eqt:DeltaPSF+-DeltaPSF-evenonZernikeMatrix}
\end{align}

where $\underline{Z}_f$ is the $N^2 \mathrm{x} k_{odd}$ matrix regrouping all the terms of $2\Im\lbrace \widetilde{P^*}\widetilde{P Z_j}\rbrace$, $k_{odd}$ is the number of odd Zernike polynomials between $j_{min}$ and $j_{max}$, and $\underline{Z}_d$ is the $N^2 \mathrm{x} k_{even}$ matrix regrouping all the terms of $\lbrace-4\widetilde{PC}^*\widetilde{SZ_j} +4\widetilde{PS}^*\widetilde{CZ_j}\rbrace$, $k_{even}$ is the number of even Zernike polynomials between $j_{min}$ and $j_{max}$.

\subsection{Implementation}

The resolution of the equations as said above is done with a linear regression. The code is structured as following :

\begin{description}
\item[The Class phaseDiversity3PSFs] is the main class of the program, see Appendix \ref{subapp:phaseDiversity3PSFs}. It takes as inputs : the 3 PSFs needed to retrieve the phase as exposed above (inFoc, outFocpos and outFocneg), the displacement of the detector with respect to its focused position ($\Delta z$),

\begin{equation}
\Delta z = \frac{4\Delta\phi}{\pi}\lambda \left(\frac{D}{F}\right)^2,
\label{eqt:Deltaz}
\end{equation}

where $\Delta\phi$ is the Peak To Valley (P2V) dephasing that we want to introduce in the defocused image in the following it will be equal to $2\pi$, $D$ is the pupil diameter and $F$ the focal length of the optical system. It takes also as inputs the wavelength of the incoming light, the pixel size of the detector, the focal length of the optical system, the radius of the pupil and the boundary on $jmin<j<jmax$ the Zernike index. 

The instantiation of all the elements of eqts. \eqref{eqt:DeltaPSFoddonZernikeMAtrix} and \eqref{eqt:DeltaPSF+-DeltaPSF-evenonZernikeMatrix} is done by calling the class method \verb!initiateMatrix1()! and \verb!initiateMatrix2()!. These two methods compute the left and right members of the equations by calling methods present in the script \verb!fs.py!.

\item[The script fs] is  gathering functions needed to retrieve the phase of the wavefront, see Appendix \ref{subapp:fs}. $\Delta PSF_{odd}$ and $\Delta PSF_{\delta\phi_+, even}-\Delta PSF_{\delta\phi_-, even}$ are computed by the two methods \verb!y1(params)! and \verb!y2even(params)!. The matrix elements of $\underline{Z}_f$ are computed by the method \verb!f1j(params)! and the elements of $\underline{Z}_d$ are computed by \verb!f2jeven(params)!. Those two functions needs the Zernike polynomial basis which is coded in \verb!zernike.py!, see Appendix \ref{subapp:zernike}.
\end{description}

The linear regression is done after having initiated all the elements of the equations. We use the \verb!numpy.linalg! class, especially its \verb!lstq! function\footnote{Documentation found at \url{https://docs.scipy.org/doc/numpy/reference/generated/numpy.linalg.lstsq.html\#numpy.linalg.lstsq}}. This functions takes a vector $\vec{b}$ corresponding to the flattened $Delta PSF_{odd}$ or $\Delta PSF_{\delta\phi_+, even}-\Delta PSF_{\delta\phi_-, even}$ and a matrix  $\underline{a}$ corresponding $\underline{Z}_f$ or $\underline{Z}_d$. Then it determines the vector $\vec{x}$ corresponding to the vector $\vec{a}$ that minimizes the euclidean 2-Norm $\lVert b-ax \rVert$ by computing the Moore-Penrose pseudo inverse of $\underline{a}$.

\section{Results}
\label{sec:ourPDresult}


